Titeltext

### 2018-07-25 thesis/paper/paper.tex
Added basic modelling and more constraints

\begin{table}
\caption{wikidata datatypes}\label{tab_datatypes}
\begin{tabularx}{\textwidth}{lllX}
Datatype & Fields & Description \\
\hline
string & string (string) & simple string literal \\
globecoordinate & \makecell{latitude (float) \\ longitude (float) \\ precision (float) \\ globe(URI)} & \makecell{coordinates on a given celestial body \\ or geographic standard} \\
quantity & \makecell {amount (decimal) \\ upperBound (decimal) \\ lowerBound (decimal) \\ unit (URI or 1)} & quantity of a specified unit (or none) \\ with uncertainty \\
time & \makecell {time (string) \\ calendarmodel (URI) \\ precision (integer)} & \makecell {time as a string according to calendarmodel \\ (julian or gregorian),\\ with precision mapping to orders of magnitude \\ (eg. 10 years, months)} \\

@@

While this is simliar to RDF and Wikidata has in fact been exported as RDF \cite{EGKMV2014}, the data structure is more complex than simple triples. It allows, in essence, to store additional information about statements. This is done via a statement id, which is assigend to every triple storing information about an item. 
\section{Datalog}
Datalog is a logic programming language. It is fully delarative and uses simple rules to derive additional knowledge from given facts. The most basic unit of information in datalog is a constant, which is a given [string]? expressing a fix point. In the context of Wikidata any item or property, e.g. the universe (Q1) or head of government (P6), could be represented as a constant. 
\section{Terminology}
\begin{table}[H]
\caption{shorthands}
\begin{tabular}{ll}
Shorthand & Description \\
\hline
constrained statement & statement with the constrained property as predicate \\
item statement & statement on the same item as the constrained statement \\
constrained item & item with a constrained statement
\end{tabular}
\end{table}
\chapter{Property Constraints}

@@

%\# of properties & constraint id & constraint name \\
%\hline
%4265 & Q21503247 & Item requires statement \\
%3021 & Q21502404 & Format \\
%2909 & Q19474404 & Single value \\
%2843 & Q21502410 & Unique value \\
%2613 & Q21503250 & Type \\
%1410 & Q53869507 & Scope \\
%781 & Q21502838 & Conflicts with \\
%707 & Q21510865 & Value type \\
%343 & Q21514353 & Allowed units \\
%303 & Q21510851 & Allowed qualifiers \\
%244 & Q21510864 & Value requires statement \\
%203 & Q21510856 & Mandatory qualifier \\
%188 & Q21510860 & Range \\
%119 & Q21510855 & Inverse \\
%116 & Q21510859 & One of \\
%106 & Q52848401 & Integer \\
%65 & Q51723761 & No bounds \\
%53 & Q21510852 & Commons link \\
%33 & Q21510862 & Symmetric \\
%27 & Q52004125 & Allowed entity types \\
%17 & Q52558054 & None of \\
%13 & Q21510857 & Multi-value \\
%10 & Q54554025 & Link \\
%6 & Q21510854 & Difference within range \\
%3 & Q52060874 & Single best value \\
%1 & Q25796498 &  \\
%1 & Q52712340 &  \\
%
%\end{tabular}
%\end{table}
To improve the data quality, help editors avoid common mistakes and clarify the usage of properties, Wikidata allows constraints to be defined on properties. They are however not firm rules and execptions can and should be made if necessary. A property constraint says something about the way a property may or may not be used. This can mean simple restrictions on the statement a property is part of, e.g. only certain values are allowed or more complex requirements on connected parts of the data graph, e.g. the value of a statement with this property should have certain statements. On the following page you can find table \ref{tab_basestats} which list the existing property constraints and a short description as well as classifying information which will be explained shortly. Afterwards the constraints will be explained in detail and either a basic approach for their modeling or reasoning why it is not possible or sensible to model it given.\\

@@

\subsection{Single value (Q19474404)}
The single value constraint constrains the values a property may take on a single item. A property marked with the single value constraint can have properties marked as separators, which define exceptions under specific circumstances.
\begin{table}[H]
\caption{single value constraint (Q19474404)}
\begin{tabularx}{\textwidth}{ ll X}
\hline
Requirement & constrained statement value \\
\hline
Qualifiers & \makecell{required property (P2306) -- 1 \\ property the item statement must have \\ allowed value (P2305) -- 0..* \\ values the item statement may have} \\
\hline
Violation & \makecell{no item statement with required property \\ all item statements with required property have no allowed value} \\
\hline
Dependency &  all statements of a constrained item\\
\hline
Rule features & negation \\
\hline
\end{tabularx}
\end{table}

\chapter{Rules}
\chapter{Evaluation/Comparison}
\end{document}

>>>>

\begin{table}[H]
\caption{wikidata datatypes}\label{tab_datatypes}
\begin{tabularx}{\textwidth}{lllX}
Datatype & Fields & Description \\
\hline
string & string (string) & simple string literal \\
globecoordinate & \makecell{latitude (float) \\ longitude (float) \\ precision (float) \\ globe(URI)} & \makecell{coordinates on a given celestial body \\ or geographic standard} \\
quantity & \makecell {amount (decimal) \\ upperBound (decimal) \\ lowerBound (decimal) \\ unit (URI or 1)} & quantity of a specified unit (or none) \\ with uncertainty \\
time & \makecell {time (string) \\ calendarmodel (URI) \\ precision (integer)} & \makecell {time as a string according to calendarmodel \\ (julian or gregorian),\\ with precision mapping to orders of magnitude \\ (eg. 10 years, months)} \\

@@

While this is simliar to RDF and Wikidata has in fact been exported as RDF \cite{EGKMV2014}, the data structure is more complex than simple triples. It allows, in essence, to store additional information about statements. This is done via a statement id, which is assigend to every triple storing information about an item. 
\section{Datalog}
Datalog is a logic programming language. It is fully delarative and uses simple rules to derive additional knowledge from given facts. The most basic unit of information in datalog is a constant, which is a given [string]? expressing a fix point. In the context of Wikidata any item or property, e.g. the universe (Q1) or head of government (P6), could be represented as a constant.
\section{Representation}
To model the basic represenation of triples, qualifiers and references three predicates are introduced.\\
tripleEDB (STATEMENT, ITEM, PROPERTY, VALUE) represents a single statement consisting of the statement ID, the item ID, the property ID and the value as either and item ID or a string-representation of the data value.\\
qualifierEDB (STATEMENT, PROPERTY, VALUE) and referenceEDB (STATEMENT, PROPERTY, VALUE) represent the qualifiers and references on a given statement ID with their respective property id and the value as usual.\\

Additional predicates modelling different preprocessing steps are explained in chapter \ref{cha_rules}.
\section{Terminology}
\begin{table}[H]
\caption{shorthands}
\begin{tabular}{ll}
Shorthand & Description \\
\hline
constrained statement & statement with the constrained property as predicate \\
item statement & statement on the same item as the constrained statement \\
constrained item & item with a constrained statement
\end{tabular}
\end{table}
\chapter{Property Constraints}

@@

%\# of properties & constraint id & constraint name \\
%\hline
%4265 & Q21503247 & Item requires statement \\
%3021 & Q21502404 & Format \\
%2909 & Q19474404 & Single value \\
%2843 & Q21502410 & Unique value \\
%2613 & Q21503250 & Type \\
%1410 & Q53869507 & Scope \\
%781 & Q21502838 & Conflicts with \\
%707 & Q21510865 & Value type \\
%343 & Q21514353 & Allowed units \\
%303 & Q21510851 & Allowed qualifiers \\
%244 & Q21510864 & Value requires statement \\
%203 & Q21510856 & Mandatory qualifier \\
%188 & Q21510860 & Range \\
%119 & Q21510855 & Inverse \\
%116 & Q21510859 & One of \\
%106 & Q52848401 & Integer \\
%65 & Q51723761 & No bounds \\
%53 & Q21510852 & Commons link \\
%33 & Q21510862 & Symmetric \\
%27 & Q52004125 & Allowed entity types \\
%17 & Q52558054 & None of \\
%13 & Q21510857 & Multi-value \\
%10 & Q54554025 & Link \\
%6 & Q21510854 & Difference within range \\
%3 & Q52060874 & Single best value \\
%1 & Q25796498 &  \\
%1 & Q52712340 &  \\
%
%\end{tabular}
%\end{table}
To improve the data quality, help editors avoid common mistakes and clarify the usage of properties, Wikidata allows constraints to be defined on properties. They are however not firm rules and execptions can and should be made if necessary. A property constraint says something about the way a property may or may not be used. This can mean simple restrictions on the statement a property is part of, e.g. only certain values are allowed or more complex requirements on connected parts of the data graph, e.g. the value of a statement with this property should have certain statements. On the following page you can find table \ref{tab_basestats} which list the existing property constraints and a short description as well as classifying information which will be explained shortly. Afterwards the constraints will be explained in detail and either a basic approach for their modelling or reasoning why it is not possible or sensible to model it given.\\

@@

\subsection{Single value (Q19474404)}
The single value constraint constrains the values a property may take on a single item. A property marked with the single value constraint can have properties marked as separators, which define exceptions under specific circumstances.
\begin{table}[H]
\caption{single value constraint (Q19474404)}
\begin{tabularx}{\textwidth}{ ll X}
\hline
Requirement & constrained statement value \\
\hline
Qualifiers & \makecell{separator (P4155) -- 0..* \\ qualifiers that must be different for exception} \\
\hline
Violation & \makecell{two constrained statements with same item and value \\ (exception: statements with separators as qualifiers with different values)} \\
\hline
Dependency & \makecell{without separators: all constrained statements on one item \\ with separators: additionally all qualifiers with separators on those statements }\\
\hline
Rule features & inequality, partially negation \\
\hline
\end{tabularx}
\end{table}

If no separators are specified, two triple form a violation of this constraint if they a) belong to the same item, b) have the same constrained property, and c) have the same value.
If separators are specified, a violation as above can be ignored if the statements each have at least one qualifier with the same separator as predicate and different values.\\
In the case without separators the modelling idea would be to find two triples tripleEDB with the same item and constrained property and test if their id is unequal. If separators were specified a violation would be the two triples and all their separator qualifiers either equal or nonexistent.

\subsection{Distinct values (Q21502410)}
The distinct values constraint forbids the existence of two constrained statements with the same value.
\begin{table}[H]
\caption{distinct values constraint (Q21502410)}
\begin{tabularx}{\textwidth}{ ll X}
\hline
Requirement & constrained statement value \\
\hline
Violation & two constrained statements with same value \\
\hline
Dependency & all constrained statements\\
\hline
Rule features & inequality \\
\hline
\end{tabularx}
\end{table}
To model the distinct values constraint it is necessary to find triples tripleEDB

\chapter{Rules}\label{cha_rules}
\chapter{Evaluation/Comparison}
\end{document}
